---
title: "Group project - Wine - Jared Martin - R Code"
author: "Jared Martin"
date: "2024-11-24"
output:
  pdf_document: default

---

```{r}

library(tidyverse)
library(class)
library(tree)
library(randomForest)
library(e1071)
library(nnet)

```

Group Members: 1 of 1 Jared Martin

# Cleaning and Preparing the data

The following code pulls in the two wine data sets, creates a new variable called wine type (red or white), makes quality and type factor variables, joins the two data sets together and saves the data as a csv.

```{r}

rbind(read.csv("./Data/winequality-red.csv", sep = ";") %>%
                     mutate(wine.type = "red"),
                   read.csv("./Data/winequality-red.csv", sep = ";") %>%
                     mutate(wine.type = "white"))  %>% 
  write.csv( "./Data/wine_data.csv")

wine_data <- read_csv("./Data/wine_data.csv")

wine_data$quality <- factor(wine_data$quality)

wine_data$wine.type <- factor(wine_data$wine.type)

wine_data[,2:14] -> wine_data

wine_data %>% 
  dplyr::select(quality, everything()) -> wine_data

```
# Research Question #1: Can we predict the quality of a wine given its characteristic?

Each wine was given as ranking for quality between 0-10. Although in the data set the lowest level of quality is 3, while the highest is an 8. Given that there is more then two level of quality we will be using KKN and classification trees to predict the quality of the wine given it's characteristics. 

```{r}

levels(wine_data$quality)

```

## Method #1: KNN

### 1. Spliting the data into a testing and training dataset. 

```{r}
set.seed(1)

n <- nrow(wine_data)

z <- sample(n, n/2)

wine_test <- wine_data[z,]

wine_train <- wine_data[-z,]

x.train <- wine_train[,2:12]
x.test <- wine_test[,2:12]

y.train <- wine_train$quality  
y.test <- wine_test$quality

dim(x.train)[1] == length(y.train)  # Test to see if length is the same. Should be TRUE
dim(x.test)[1] == length(y.test)    # Test to see if length is the same. Should be TRUE

```

### 2. Determining the optimal number of nearest neighbors "K"

```{r}

class.rate = rep(0,20)      

for (K in 1:20) {
  knn.result = knn(x.train, x.test, y.train, K)
  class.rate[K] = mean( y.test == knn.result )
}

plot(class.rate)

max(class.rate) # optimal number of K neighbors is 1

```
### 3. Cross Validation - CClasification Rate and Confusion table using Optimal number of nearest Neighbors. 

```{r}

knn_cv_wine_quality <- knn(x.train, x.test, y.train, 1) 

table(y.test, knn_cv_wine_quality) 

knn_cv_classification_rate <- mean(y.test == knn_cv_wine_quality)

# Classification rate of .766

```
### 4. LOOCV - CClasification Rate and Confusion table using Optimal number of nearest Neighbors.

```{r}

y.test <- wine_data$quality

x.train <- wine_data[,2:12]

knn_loocv_wine_quality <- knn.cv(x.train, y.test, k = 1, prob = FALSE)

table(y.test, knn_loocv_wine_quality)

knn_loocv_classification_rate <- mean(y.test == knn_loocv_wine_quality)

#Wow! Classification rate of 1!

```
## Method #2: Classification Trees

### 1. Classification Tree for Quality 

```{r}

tree_wine_quality <- tree(quality ~ ., data = wine_train)

tree_wine_quality

summary(tree_wine_quality)

# not all class are represented as terminal node classifications (3, 4 & 8 are not represented). 

```
### 2. Calculating the Classification Rate for Unprunned Tree 

```{r}

tree_hat <- predict(tree_wine_quality, wine_test, type = "class")

table(tree_hat, wine_test$quality)

tree_classification_rate <- mean(tree_hat == wine_test$quality)

```

### 3. Prunning the Classification tree on misclassifaction rate. 

```{r}

# The optimal number of terminal nods on a tree is 8. The same number of terminal nods as the tree above; therefore no pruning is required. 

set.seed(1)

tree_cv_wine_quality <- cv.tree( tree_wine_quality, FUN = prune.misclass )  

tree_cv_wine_quality

plot(tree_cv_wine_quality)

tree_pruned_wine_quality <- prune.misclass(tree_wine_quality, best = 6)

```
### 4. Plotting Classification Tree

```{r}

plot(tree_pruned_wine_quality, type = "uniform")
text(tree_pruned_wine_quality, cex = 0.6)

```
### 5. Cross Validation of the Classification Tree

```{r}

tree_hat <- predict(tree_pruned_wine_quality, wine_test, type = "class")

table(tree_hat, wine_test$quality)

tree_pruned_classification_rate <- mean(tree_hat == wine_test$quality)

```
##Method #3: Random Forest

### 1. Creaing the Traing Random Forest

```{r}

set.seed(1)

forest_wine_quality <- randomForest(quality ~ ., data = wine_train)

forest_wine_quality

importance(forest_wine_quality)

varImpPlot(forest_wine_quality)

```
### 2. Searching for the optimal number of trees and number of variables considerd at each branch.

```{r}

set.seed(1)

err.rate <- numeric(10)  # To store error rates for each mtry
n.trees <- numeric(10)   # To store the number of optimal trees for each mtry

for (m in 1:10) {
  forest_wine_quality_m <- randomForest(quality ~ ., data = wine_train, 
                                  mtry = m )
  opt.tree <- which.min(forest_wine_quality_m$err.rate)
  forest_wine_quality_m <- randomForest(quality ~ ., data = wine_train,
                                        mtry = m, ntree = opt.tree)
  yhat_rf_m <- predict(forest_wine_quality_m, newdata = wine_test, type = "class")
  err.rate[m] <- 1-mean(wine_test$quality == yhat_rf_m)
  n.trees[m] <- opt.tree
}

err.rate
n.trees

plot(err.rate); line(err.rate)

plot(n.trees); line(n.trees)

# optimal solution is 501 trees and 8 variables considered at each branch. 

```
### 3. Cross Validation with Optimal number of trees and number of variables considerd at each branch.

```{r}

set.seed(1)

forest_cv_wine_quality <- randomForest(quality ~ ., data = wine_train, 
                                  mtry = 8, ntree = 501 )

yhat_rf <- predict(forest_cv_wine_quality, newdata = wine_test, type = "class")

forest_classification_rate <- mean(wine_test$quality == yhat_rf)

```

## Method #4: Support Vecot Machines

### 1. Creating a SVM model with a linear bounder

```{r}

SVM_wine_quality <- svm(quality ~ ., data = wine_train, kernel="linear")

summary(SVM_wine_quality)

```
### 2. Tuning the SVM model to detiermen the optimal kernel and cost. 

```{r}

set.seed(1)

svm_tuned <- tune(
  svm,
  quality ~ ., 
  data = wine_train,
  ranges = list(
    cost = 10^seq(-2, 2, by = 1),  
    kernel = c("linear", "polynomial", "radial", "sigmoid")  # Test multiple kernels
  )
)

```

```{r}

summary(svm_tuned)

# optimal cost is 10 with a radial boundary. 

```
### 3. Cross Validation with Optimal Cost and Kernal 

```{r}

SVM_wine_quality_optimal <- svm(quality ~ ., 
                                data = wine_train, 
                                kernel="radial",
                                cost = 100)

summary(SVM_wine_quality_optimal)

svm_yhat <- predict(SVM_wine_quality_optimal, data = wine_test)

table(svm_yhat, wine_test$quality)

svm_cv_classification_rate <- mean(svm_yhat == wine_test$quality)

```
## Method #5: Artifical Neural Nextwork

### 1. Searching for the optimal size of hiden nodes

```{r}

set.seed(1)

err.rate <- numeric(10)  

for (m in 1:12) {
  ann_wine_quality <- nnet(quality ~ ., 
                         data = wine_train,
                         size = m)
  yhat_ann <- predict(ann_wine_quality, newdata = wine_test, type = "class")
  err.rate[m] <- 1-mean(wine_test$quality == yhat_ann)
}

err.rate

plot(err.rate)

# 11 hidden nodes produce the lowest prediction error rate

```
### 2. Cross Validation with Optimal number of hidden nodes

```{r}

set.seed(1)

ann_wine_quality_optimal <- nnet(quality ~ ., 
                         data = wine_train,
                         size = 11)

yhat_ann <- predict(ann_wine_quality, newdata = wine_test, type = "class")
  
ann_classification_rate <- mean(wine_test$quality == yhat_ann)

```


## Summary: Which model preformed at predictiong the quality of a wine given it's chariterisitcs? 

Using cross validation evaluating the performance of each model, the random forest model with 501 trees and 8 mrty produced the lowest error rate of 15.2%. 

```{r}

classification_rates <- data.frame(
  Model = c("K-Nearest Neighbors", "Classifcation Tree", "Random Forest", "Support Vector Machines", "Artifical Nerual Network"),
  Classification_Rate = c(knn_cv_classification_rate,
                         tree_pruned_classification_rate,
                         forest_classification_rate,
                         svm_cv_classification_rate, 
                         ann_classification_rate),
  Error_Rate = c(1- knn_cv_classification_rate,
                      1- tree_pruned_classification_rate,
                      1- forest_classification_rate,
                      1- svm_cv_classification_rate,
                      1- ann_classification_rate)
)

classification_rates %>% 
  arrange(desc(Classification_Rate))

```

