---
title: "Group project - Wine - Jared Martin - R Code"
author: "Jared Martin"
date: "2024-11-24"
output:
  pdf_document: default

---

```{r}

library(tidyverse)
library(class)
library(tree)

```

# Cleaning and Preparing the data

The following code pulls in the two wine data sets, creates a new variable called wine type (red or white), makes quality and type factor variables, joins the two data sets together and saves the data as a csv.

```{r}

rbind(read.csv("./Data/winequality-red.csv", sep = ";") %>%
                     mutate(wine.type = "red"),
                   read.csv("./Data/winequality-red.csv", sep = ";") %>%
                     mutate(wine.type = "white"))  %>% 
  write.csv( "./Data/wine_data.csv")

wine_data <- read_csv("./Data/wine_data.csv")

wine_data$quality <- factor(wine_data$quality)

wine_data$wine.type <- factor(wine_data$wine.type)

wine_data[,2:14] -> wine_data

wine_data %>% 
  dplyr::select(quality, everything()) -> wine_data

```
# Research Question #1: Can we predict the quality of a wine given its characteristic?

Each wine was given as ranking for quality between 0-10. Although in the data set the lowest level of quality is 3, while the highest is an 8. Given that there is more then two level of quality we will be using KKN and classification trees to predict the quality of the wine given it's characteristics. 

```{r}

levels(wine_data$quality)

```

## Method #1: KNN

### 1. Spliting the data into a testing and training dataset. 

```{r}
set.seed(1)

n <- nrow(wine_data)

z <- sample(n, n/2)

wine_test <- wine_data[z,]

wine_train <- wine_data[-z,]

x.train <- wine_train[,2:12]
x.test <- wine_test[,2:12]

y.train <- wine_train$quality  
y.test <- wine_test$quality

dim(x.train)[1] == length(y.train)  # Test to see if length is the same. Should be TRUE
dim(x.test)[1] == length(y.test)    # Test to see if length is the same. Should be TRUE

```

### 2. Determining the optimal number of nearest neighbors "K"

```{r}

class.rate = rep(0,20)      

for (K in 1:20) {
  knn.result = knn(x.train, x.test, y.train, K)
  class.rate[K] = mean( y.test == knn.result )
}

plot(class.rate)

max(class.rate) # optimal number of K neighbors is 1

```
### 4. Cross Validation - CClasification Rate and Confusion table using Optimal number of nearest Neighbors. 

```{r}

knn_cv_wine_quality <- knn(x.train, x.test, y.train, 1) 

table(y.test, knn_cv_wine_quality) 

knn_cv_classification_rate <- mean(y.test == knn_cv_wine_quality)

# Classification rate of .766

```
### 3. LOOCV - CClasification Rate and Confusion table using Optimal number of nearest Neighbors.

```{r}

y.test <- wine_data$quality

x.train <- wine_data[,2:12]

knn_loocv_wine_quality <- knn.cv(x.train, y.test, k = 1, prob = FALSE)

table(y.test, knn_loocv_wine_quality)

knn_cv_classification_rate <- mean(y.test == knn_loocv_wine_quality)

#Wow! Classification rate of 1!

```
## Method #2: Classification Trees

### 1. Classification Tree for Quality 

```{r}

tree_wine_quality <- tree(quality ~ ., data = wine_train)

tree_wine_quality

summary(tree_wine_quality)

# not all class are represented as terminal node classifications (3, 4 & 8 are not represented). 

```
### 2. Calculating the Classification Rate for Unprunned Tree 

```{r}

tree_hat <- predict(tree_wine_quality, wine_test, type = "class")

table(tree_hat, wine_test$quality)

tree_classification_rate <- mean(tree_hat == wine_test$quality)

```

### 3. Prunning the Classification tree on misclassifaction rate. 

```{r}

# The optimal number of terminal nods on a tree is 8. The same number of terminal nods as the tree above; therefore no pruning is required. 

set.seed(1)

tree_cv_wine_quality <- cv.tree( tree_wine_quality, FUN = prune.misclass )  

tree_cv_wine_quality

plot(tree_cv_wine_quality)

Tree_pruned_wine_quality <- prune.misclass(tree_wine_quality, best = 6)

```
### 4. Plotting Classification Tree

```{r}

plot(Tree_pruned_wine_quality, type = "uniform")
text(Tree_pruned_wine_quality, cex = 0.6)

```
### 5. Cross Validation of the Classification Tree

```{r}

tree_hat <- predict(Tree_pruned_wine_quality, wine_test, type = "class")

table(tree_hat, wine_test$quality)

tree_pruned_classification_rate <- mean(tree_hat == wine_test$quality)

```

