---
title: "Group project - Wine - Jared Martin - R Code"
author: "Jared Martin"
date: "2024-11-24"
output: html_document
---

```{r}

library(tidyverse)
library(stats)

```

## Cleaning and Preparing the data

The following code pulls in the two wine data sets, creates a new variable called wine type (red or white), makes quality and type factor variables and join the two data sets togeather.

```{r}

wine_data <- rbind(read.csv("./Data/winequality-red.csv", sep = ";") %>%
                     mutate(wine.type = "red"),
                   read.csv("./Data/winequality-red.csv", sep = ";") %>%
                     mutate(wine.type = "white")) %>% 
  mutate(wine.type = as.factor(wine.type),
         quality = as.factor(quality),
         pH = as.double(pH))

sapply(wine_data, class)

```

## Research Question #1: Can we predict the type of wine given its characteristic?

### Method #1: Logistic Regression

#### Using Variable Selection: Methods to detirem the best combination of variables to use.  

```{r}

levels(wine_data$wine.type)

logistic_type <- glm( wine.type ~ pH , data = wine_data , family = "binomial" )

summary(logistic_type)

```

